A thermodynamic view of dusty protoplanetary disks I: axisymmetric stability

### 2016-11-14 paper.tex
home/work sync

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This is analogous to entropy conservation
  following an adiabatic gas.  
  Motivated by this similarity, we develop a thermodynamical
  framework to study dusty gas dynamics in protoplanetary disks. We  
 show that an isothermal dusty gas behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to deduce that   
\begin{inparaenum}[1)] 
\item 
 the settling of perfectly coupled dust cannot cause axisymmetric 
 instabilties, however thin the dust layer becomes; 
\item 
  the streaming instability neccessarily involves a pressure response
  that lags dust density; 
\item dust-loading generally stabilizes the vertical shear
  instability, because dust-gas coupling introduces buoyancy. 
\end{inparaenum}  
%growth rates, and 
%preferentially stabilize modes with small radial 
%wavelengths, as well as `surface modes' concentrated near the
%disk boundaries. 
%These results can be understood in terms of 
%buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We also discuss future applications of the dusty/adiabatic gas
correspondence in finding dusty analogs of other hydrodynamic
phenomena (e.g. Rossby wave instability, convective overstability, and 
zombie vortices), and to simulate dusty protoplanetary disks
using purely hydrodynamic codes.

@@

This paper is organized as follows. We first
explain in \S\ref{grow_osc} a general result of fluid dynamics that
underpins the rest of our study. We describe how 
a pressure-density phase lag in fluids can lead to 
instability.

This motives us to find an analogy between dusty gas and
pure hydrodynamics, as it allows us to understand dusty-gas dynamics
within a thermodynamic framework.

We begin our formal discussion in \S\ref{setup} by listing the 
one-fluid equations describing tightly-coupled dusty gas. We show that for a
fixed equation of state, these are in 
fact the usual hydrodynamic equations in disguise. 
We describe equilibrium solutions in \S\ref{eqm} that are used in
explicit calculations. In \S\ref{limits} we discuss 
general stability properties of dusty disks. We then move onto
numerical stability analyses. In  
\S\ref{linear_problem} we list the linearized equations, and solve 
them for the SI in \S\ref{si}, and in 
 \S\ref{results} to study the dusty VSI. We 
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.

>>>>

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This is analogous to entropy conservation
  following an adiabatic gas.  
  Motivated by this similarity, we develop a thermodynamical
  framework to study dusty gas dynamics in protoplanetary disks. We  
 show that an isothermal dusty gas behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to deduce that   
\begin{inparaenum}[1)] 
\item 
 the settling of perfectly coupled dust cannot cause axisymmetric 
 instabilties, however thin the dust layer becomes; 
\item 
  the streaming instability neccessarily involves a pressure response
  that lags dust density; 
\item dust-loading generally stabilizes the vertical shear
  instability, because dust-gas coupling introduces buoyancy. 
\end{inparaenum}  
%growth rates, and 
%preferentially stabilize modes with small radial 
%wavelengths, as well as `surface modes' concentrated near the
%disk boundaries. 
%These results can be understood in terms of 
%buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We also discuss future applications of the dusty/adiabatic gas
correspondence in finding dusty analogs of other hydrodynamic
phenomena (e.g. Rossby wave instability, convective overstability, and 
zombie vortices), and to simulate dusty protoplanetary disks
using purely hydrodynamic codes. %We briefly describe extending this
%equivalence principle to polytropic and adiabatic gas.

@@

This paper is organized as follows. We first
explain in \S\ref{grow_osc} a general result of fluid dynamics that
underpins the rest of our study. We describe how 
a pressure-density phase lag in fluids can lead to 
instability. Thus finding an analogy between dusty gas and
pure hydrodynamics allow us to interpret dusty-gas dynamics
within a thermodynamics framework.

We begin our formal discussion in \S\ref{setup} by listing the 
basic equations describing tightly-coupled dusty gas. We show that for a
fixed equation of state, these are in 
fact the usual hydrodynamic equations in disguise. 
We describe equilibrium solutions in \S\ref{eqm} that are used in
explicit calculations. In \S\ref{limits} we discuss 
general stability properties of dusty disks. We then move onto
numerical stability analyses. In  
\S\ref{linear_problem} we list the linearized equations, and solve 
them for the SI in \S\ref{si}, and in 
 \S\ref{results} to study the dusty VSI. We 
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.
